---
title: "alameda older adult homelessness"
author: "Harrison Linder"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(ggplot2)
library(reshape2)
```
## data sources
I am using two main data sources. The first is PIT and HIC count data from HUD. This includes statistics collected during the annual Point in Time (PIT) counts and the annual counts of shelter and permanent housing available in each Continuum of Care (CoC). The specific PIT data table used was created in excel by me using excel data from HUD's website. The second data sources comes from California's HDIS data and is directly queried from their website. I look at 5 years of data from 2019 to 2023. At the time of this analysis, the 2023 data was the latest available.
```{r reading data}


HDISdata <- read.csv("https://data.ca.gov/dataset/c7ed1ae4-4f93-4fc7-b603-3cd07a55d862/resource/b1a5ae24-5842-425c-b56c-aa90f8f1c767/download/homelessness-count-by-age.csv")

HDISdata$EXPERIENCING_HOMELESSNESS_CNT <- as.numeric(HDISdata$EXPERIENCING_HOMELESSNESS_CNT) # making the count a numeric vector instead of a character vector
```
```{r}
ggplot(HDISdata[HDISdata$LOCATION_ID == "CA-502" & HDISdata$CALENDAR_YEAR != "2024",], aes(x = factor(CALENDAR_YEAR), y = EXPERIENCING_HOMELESSNESS_CNT, color = AGE_GROUP_PUBLIC, group = AGE_GROUP_PUBLIC)) +
  geom_line(size = 1) +  # Draw the lines
  geom_point(size = 2) +  # Add points to the lines
  labs(
    x = "Year",
    y = "Alameda HDIS Homeless Count by Age",
    title = "Alameda Homeless Population Counts Over Time"
  )
```
